---
title: "Data Wrangling (Data Preprocessing)"
author: "Farhan Ahmed"
subtitle: Practical assessment 2
date: "07/10/2022"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
editor_options: 
  markdown: 
    wrap: 72
---

## **Setup**

```{r}

# Load the necessary packages required to reproduce the report. For example:

library(dplyr)
library(kableExtra)
library(magrittr)
library(readr)
library(tidyr)
library(readxl)
library(outliers)
```

## **Student names, numbers and percentage of contributions**

```{r, echo=FALSE}

# Add your names, numbers and percentage of your contribution here.

na<- c("Farhan Ahmed","Jack Mulder")
no<- c("s3914456","s3954768")
pc<- c("50%","50%")

s<- data.frame(cbind(na,no,pc))
colnames(s)<- c("Student name", "Student number", "Percentage of contribution")

s %>% kbl(caption = "Group information") %>%
  kable_classic(full_width = F, html_font = "Cambria")

```

<br> <br>

## **Executive Summary**

The first step in processing the data was tidying the untidy dataset. This was done by making sure all variables had one column and each row being a single observation this was achieved through using pivot_longer. The two data sets where then merged.  Once merged two new columns were created one being the ratio of household income to college tuition, the other being a factor ordering household income status as either low, medium, or high.
All numeric columns bar year were inspected for outlier using either Tukey’s method or z-score depending on whether the data was normalized or not according to the Shapiro-Wilk test for normalization. Outlier where only detected in the non-normalized data, this meant that a transformation could be used to handle any outliers. 
Histograms were used to determine whether data was left or right skewed, once they were identified an appropriate data transformation could take place, this was log10 or reciprocal for right skewness and a cubed transformation for left skewed data. Then z-scores where used to confirm that their were no outliers and the Shapiro-Wilk test to confirm that the data was normally distributed. 


<br> <br>

## **Data**

\<!--\> What data is included, and what does it represent? What is the
source of the data? How was it collected?\<--\>

We have used two data sets for our project. One of them is an untidy
data set and the other one is a tidy data set. The first data set used
was, US states with average tuition fees from the year 2004 to 2016
(xlsx file format). This was an untidy data set. This data set was
collected from gitHub(us_avg_tuition.xlsx, 2022). The second data set,
US state demographics for 2014, was collected from github(CORGIS Datasets Project, 2022) but was
originally collected from United States Census Bureau.

-   tidy data set - US state demographics with 48 variables with 51
    observations
-   In this data set 6 of the variables were focused on
    -   State is a string variable containing the name of the sate
    -   Population.2014.Population is a numeric variable, containing the
        population of each state in 2014
    -   Education.Bachelor.s.Degree.or.Higher is numeric variable,
        containing the percentage of the population who have a bachelors
        degree or higher
    -   Education.High.School.or.Higher is an numeric variable,
        containing the percentage of the population who have completed
        high school or higher
    -   Income.Per.Capita.Income is a numeric variable, containing the
        income per capita for the state in USD
    -   Income.Median.Houseold.Income is a numeric variable, containing
        the median household income for each state in USD
-   Untidy data set - US state average college tuition
    -   Each observation had a variable state, which is a string
        containing the state name, as well as the cost of tuition for
        each year in USD as numeric variable

```{r}
# Reading the xlsx file and storing it
tuition_data <- read_excel("us_avg_tuition.xlsx")

# Displaying the first 6 observations
head(tuition_data)
```

-   Tidy data set - US state-wise demographics (csv file format)

```{r}
# Reading the csv file and storing it
state_data<-read.csv("state_demographics.csv")

# Selecting specific variables and storing them
state_data<-state_data%>%select(State, Population.2014.Population, Education.Bachelor.s.Degree.or.Higher, Education.High.School.or.Higher,Income.Per.Capita.Income, Income.Median.Houseold.Income)

# Displaying the first 6 observations
head(state_data)
```

Before we go ahead and merge the two data sets based on the common key
variable, we need to first convert the untidy data set into proper
structure using the tidy principles.

<br> <br>

## **Tidy & Manipulate Data I**

The US average tuition fees data set is untidy because the column
headers except the first column 'State' are all values of a variable and
not variable names.

-   The 12 columns from 2004-05 all the way to 2015-16 should represent
    as values of the year variable.

-   The values of the year variable should be separated such as
    (2004,2005,..)

So this way we could use the apply tidy principles by each variable
having one column, each row having one observation and each value having
one cell.

```{r}
# Applying wide to long formatting from column 2 till column 13
tuition_data <- tuition_data %>% pivot_longer(names_to = "year", values_to = "Avg tuition (USD)", cols = 2:13)
```

Here we convert from wide to long format by using the function
pivot_longer() which is an updated alternate to gather() function. We
specify the names_to as the name of our variable 'year' containing the
years as values and values_to as 'Avg tuition (USD)' containing the
tuition fee values.

```{r}
# Separating year to achieve one row one observation
tuition_data <- tuition_data %>% separate(year, into = c("year"), sep ="-")
# Converting type of variable year to double
tuition_data$year %<>% as.double()
```

We are now separating the year variable to keep the year column
structured as YYYY, therefore 2004-05 is now 2004, 2005-06 is now 2005
and so on.

<br> <br>

## **Tidy & Manipulate Data II**

We will be focusing on the US states demographic data set here, since
the demographic statistics of the data are of the year 2014, we will
first add a variable year containing value 2014.

```{r}
# Since we will be focusing only for the year 2014, we add a variable year to the other data set
state_data$year<-2014
```



```{r}
# Merging the two data sets using full join
merged_data<-full_join(tuition_data, state_data)

# Using filter() to extract and view observations only for the year 2014
merged_data%<>% filter(year==2014)

# Displaying the first 6 observations
head(merged_data)
```



We will be focusing only on year 2014 as the state demographics of the
data set contains data from the year 2014, so it would make more sense
focusing on this year only.

Both the data sets have been merged and stored as 'merged_data' and we
have used the filter() function from the dplyr library to extract
observations belonging to the year 2014.
```{r}
merged_data%<>% mutate(Income.status = case_when(Income.Median.Houseold.Income<60000~"Low",
                                              Income.Median.Houseold.Income>60000 & Income.Median.Houseold.Income<72500~"Medium",
                                              Income.Median.Houseold.Income>72500~"High"))
```

```{r}
merged_data<- mutate(merged_data, income.to.tuition.cost.ratio = `Income.Median.Houseold.Income`/`Avg tuition (USD)`)
```

Since both are data sets are in the appropriate structure, we will now
be merging the two data sets based on the common key variable 'State'
and 'year
<br> <br>

## **Understand**

```{r}
# This is the R chunk for the Tidy & Manipulate Data I 

```

Provide explanations here.

<br> <br>

## **Scan I**

```{r}
# This is the R chunk for the Scan I

```

Provide explanations here.

<br> <br>

## **Scan II**

```{r}
par(mfrow = c(2, 2)) 
merged_data$income.to.tuition.cost.ratio %>% 
  boxplot(main = "Box Plot of income to tuition ratio", 
          ylab = "ratio")

merged_data$`Avg tuition (USD)` %>% 
  boxplot(main = "Box Plot of averge tuition cost", 
          ylab = "tuition cost(USD)") 

merged_data$Income.Median.Houseold.Income%>% 
  boxplot(main = "Box Plot of Median household incomes", 
          ylab = "$USD") 

merged_data$Education.Bachelor.s.Degree.or.Higher %>% 
  boxplot(main = "Box Plot %bachelors degree", 
          ylab = "% of the population")
par(mfrow = c(1, 1))


```
```{r}

par(mfrow = c(2, 2))

merged_data$Education.High.School.or.Higher %>% 
  boxplot(main = "Box Plot  of % High school degree", 
          ylab = "% of the population")
merged_data$Income.Per.Capita.Income %>% 
  boxplot(main = "Box Plot of Income per Capita ", 
          ylab = "$USD")
merged_data$Population.2014.Population %>% 
  boxplot(main = "Population in 2014", 
          ylab = "Population 1000")
par(mfrow = c(2, 2))
```
Using box plots we can scan for any outliers within the data. However this method is useful it only works on non-normal data, so it should be checked if the data is normally distributed.  

```{r}
shapiro.test(merged_data$income.to.tuition.cost.ratio)
shapiro.test(merged_data$`Avg tuition (USD)`)
shapiro.test(merged_data$Income.Median.Houseold.Income)
shapiro.test(merged_data$Education.Bachelor.s.Degree.or.Higher)
shapiro.test(merged_data$Education.High.School.or.Higher)
shapiro.test(merged_data$Income.Per.Capita.Income)
shapiro.test(merged_data$Population.2014.Population)


```
Using the Shapiro Wilk test it can be seen that two of columns are normally distributed, as the p-value score is greater than 0.05 meaning that that the null hypothesis that the data is normally distributed is accepted. This means Tukeys method should not be used on average tuition and median household income as they are normally distributed.

With the columns that are not normally distributed, they will first be transformed to decrease the skweness and convert to a normal distribution and to get rid of any outliers.

```{r}
z.score<- merged_data$`Avg tuition (USD)`%>% scores(type = "z")
length(which(abs(z.score) >3 ))
z.score1<- merged_data$Income.Median.Houseold.Income%>% scores(type = "z")
length(which(abs(z.score1) >3 ))



```
Here z scores have been used to detect outliers in the normally distributed data. A z score greater than 3 indicates an outlier but, here it can be seen that no outliers have been found.



## **Transform**

```{r}
par(mfrow= c(2,3))

hist(merged_data$income.to.tuition.cost.ratio, main = "Histogram of ratio")

hist(merged_data$Income.Per.Capita.Income, main = "Histogram of Ang Tuition")

hist(merged_data$Population.2014.Population, main = "Histogram of pop")

hist(merged_data$Education.Bachelor.s.Degree.or.Higher, main = "Histogram of %Bachelors")

hist(merged_data$Education.High.School.or.Higher, main = "Histogram of %Highschool")

par(mfrow= c(4,4))

```

Here we can see that that all of this data is skewed which is leading to outliers. This can also be visualized through a Q-Q plot as seen below with the population data.The deviation of the data from the red line shows that the data skewed.

```{r}
qqnorm(merged_data$Population.2014.Population, main = "Q-Q Plot of Population") 
qqline(merged_data$Population.2014.Population, col = "red", lwd = 1, lty = 2)

```
The deviation of the data from the red line shows that the data skewed.

```{r}
merged_data$Population.2014.Population <- log10(merged_data$Population.2014.Population)
merged_data$Education.Bachelor.s.Degree.or.Higher<- -1/(merged_data$Education.Bachelor.s.Degree.or.Higher)
merged_data$income.to.tuition.cost.ratio<- log10(merged_data$income.to.tuition.cost.ratio)
merged_data$Income.Per.Capita.Income<- -1/(merged_data$Income.Per.Capita.Income)

merged_data$Education.High.School.or.Higher<- merged_data$Education.High.School.or.Higher^3

shapiro.test(merged_data$income.to.tuition.cost.ratio)
shapiro.test(merged_data$Education.Bachelor.s.Degree.or.Higher)
shapiro.test(merged_data$Income.Per.Capita.Income)
shapiro.test(merged_data$Population.2014.Population)
shapiro.test(merged_data$Education.High.School.or.Higher)

z.score<- merged_data$Population.2014.Population%>% scores(type = "z")
length(which(abs(z.score) >3 ))
z.score1<- merged_data$Education.Bachelor.s.Degree.or.Higher%>% scores(type = "z")
length(which(abs(z.score1) >3 ))
z.score2<- merged_data$income.to.tuition.cost.ratio%>% scores(type = "z")
length(which(abs(z.score2) >3 ))
z.score3<- merged_data$Income.Per.Capita.Income%>% scores(type = "z")
length(which(abs(z.score3) >3 ))
z.score4<- merged_data$Education.High.School.or.Higher%>% scores(type = "z")
length(which(abs(z.score4) >3 ))
```

The data here was all not normally distributed before but is now normally distributed, this process has also removed any outliers from the data. For the transformation, the right skewed data underwent either a log10 transformation or a reciprocal transformation. As for the left skewed data it underwent a cubed transformation. 


##References
Corgis-edu.github.io. 2022. CORGIS Datasets Project. [online] Available at: <https://corgis-edu.github.io/corgis/csv/state_demographics/> [Accessed 7 October 2022].

GitHub. 2022. us_avg_tuition.xlsx. [online] Available at: <https://github.com/rfordatascience/tidytuesday/tree/master/data/2018/2018-04-02> [Accessed 7 October 2022].





